---
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{fontspec}
  - \usepackage{unicode-math}
---


```{r}
# Loading Packages
library(tidyverse)
library(readxl)
library(ggthemes)
library(skimr)
library(corrplot)
library(caret)
library(car)
library(FSA)
library(RVAideMemoire)
```

```{r}
# Importing the data set
red_wine <- read_excel("winequality-red.xlsx")
white_wine <- read_excel("winequality-white.xlsx")
```


```{r}
# Viewing the first 10 rows of the red wine data frame.
head(red_wine,10)
```
```{r}
# Viewing the first 10 rows of the white wine data frame.
head(white_wine,10)
```
```{r}
# Structure of both data frames.
str(red_wine)
str(white_wine)
```

```{r}
# checking for missing values
sum(is.na(red_wine))
sum(is.na(white_wine))
```


```{r}
# Checking for duplicates
sum(duplicated(red_wine))
sum(duplicated(white_wine)) 
```
```{r}
# Removing duplicates
red_wine <- red_wine[!duplicated(red_wine), ]
white_wine <- white_wine[!duplicated(white_wine), ]
```

```{r}
# Structure of both data frames.
str(red_wine)
str(white_wine)
```


They both have the same column names and column length.
```{r}
# Merging the data frames for easier analysis
red_wine <- mutate(red_wine, wine_type = as.factor("Red"))
white_wine <- mutate(white_wine, wine_type = as.factor("White"))
wine_data <- bind_rows(red_wine, white_wine)
```

```{r}
#converting quality to ordinal categorical
wine_data$quality <- factor(wine_data$quality, ordered = TRUE)
```


```{r}
#Checking the head and tail of the newly merged dataframe
head(wine_data,5)
tail(wine_data,5)
```
```{r}
# checking the structure of the wine_data data frame
str(wine_data)
```
The red wine had 1359 rows and 12 columns
The white wine had 3961 rows and 12 columns
The new data frame (wine_data) has 5320 row and 13 columns.
The quality column is now a categorical variable with 7 levels from 3-9
The last column represents the wine type, red or white.

```{r}
# Summary statistics
summary(wine_data)
```
From the summary statistics:
There are 30 observations with a rating of 3
There are 206 observations with a rating of 4
There are 1752 observations with a rating of 5
There are 2323 observations with a rating of 6
There are 856 observations with a rating of 7
There are 148 observations with a rating of 8
There are 5 observations with a rating of 9

```{r}
# Veiwing the summary statistics (mean and median) based on wine type
summary_stats <- wine_data %>%
  group_by(wine_type) %>%
  summarise(across(where(is.numeric), list(
    mean = mean,
    median = median
  ), .names = "{.col}_{.fn}"))
print(summary_stats)
```


```{r}
# summary statistics using the skim function
wine_data %>%
  group_by(wine_type) %>%
  skim()
```

```{r}
# Percentage distribution of quality based on the wine types
quality_percentage <- wine_data %>%
  group_by(wine_type, quality) %>%
  summarize(count = n(), .groups = "drop") %>% 
  group_by(wine_type) %>% 
  mutate(percentage = count / sum(count) * 100)
quality_percentage
```

```{r}
# Plotting the percentage distribution
ggplot(quality_percentage, aes(x = quality, y = percentage, fill = wine_type)) +
  geom_col(position = "dodge") +
  labs(title = "Percentage Distribution of Wine Quality by Type") +
  scale_fill_manual(values = c("Red" = "red", "White" = "lightblue")) +  theme_calc()
```

```{r}
# creating a function to plot box plot by wine type
plot_boxplot <- function(data, column_name) {
  ggplot(data, aes(x = wine_type, y = .data[[column_name]], fill = wine_type)) +
    geom_boxplot() +
    labs(title = paste("Boxplot of", column_name, "by Wine Type")) +
    scale_fill_manual(values = c("Red" = "red", "White" = "lightblue")) +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
}
```

```{r}
# creating a function to plot box plot by quality and wine type
plot_boxplot_quality <- function(data, column_name) {
  ggplot(data, aes(x = quality, y = .data[[column_name]], fill = wine_type)) +
    geom_boxplot() +
    labs(title = paste("Boxplot of", column_name, "by quality and Wine Type")) +
    scale_fill_manual(values = c("Red" = "red", "White" = "lightblue")) +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
}
```


```{r}
# creating a function to plot density plot
plot_density_plot <- function(data, column_name) {
  ggplot(data, aes(x = .data[[column_name]], fill = wine_type)) +
    geom_density(alpha = 0.7) +
    labs(title = paste("Density plot of", column_name, "by Wine Type")) +
    scale_fill_manual(values = c("Red" = "red", "White" = "lightblue")) +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
}
```

```{r}
# Function to perform normality test.
# Creating a function to plot QQ plot and perform Shapiro-Wilks test
normality_tester <- function(data, column_name, color, wine_type) {
  column_data <- data[[column_name]]
  
  # QQ plot
  qqplot <- ggplot(data, aes(sample = column_data)) +
    stat_qq(color = color) +
    stat_qq_line(color = "black") +
    labs(
      title = paste("QQ Plot for", column_name, "(", wine_type, ")"),
      x = "Theoretical",
      y = "Sample"
    ) +
    theme_calc()+ 
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
  print(qqplot)  
  
  #Shapiro-Wilks test
  shapiro_result <- shapiro.test(column_data)
  return(shapiro_result)
}
```

```{r}
#fixed acidity
plot_density_plot(wine_data, 'fixed acidity')
plot_boxplot(wine_data, 'fixed acidity')
plot_boxplot_quality(wine_data, 'fixed acidity')
```
```{r}
#performing normality check for red wine (fixed acidity)
normality_tester(red_wine, 'fixed acidity', 'red','Red Wine')

#performing normality check for white wine (fixed acidity)
normality_tester(white_wine, 'fixed acidity', 'lightblue', "White Wine")
```
From the normality rest, the P value is less than 0.05, therefore it doesn't follow a normal distribution


```{r}
# volatile acidity
plot_density_plot(wine_data, 'volatile acidity')
plot_boxplot(wine_data, 'volatile acidity')
plot_boxplot_quality(wine_data, 'volatile acidity')
```

```{r}
#performing normality check for red wine (volatile acidity)
normality_tester(red_wine, 'volatile acidity', 'red','Red Wine')

#performing normality check for white wine (volatile acidity)
normality_tester(white_wine, 'volatile acidity', 'lightblue', "White Wine")
```

```{r}
# citric acid
plot_density_plot(wine_data, 'citric acid')
plot_boxplot(wine_data, 'citric acid')
plot_boxplot_quality(wine_data, 'citric acid')
```

```{r}
#performing normality check for red wine (citric acid)
normality_tester(red_wine, 'citric acid', 'red','Red Wine')

#performing normality check for white wine (citric acid)
normality_tester(white_wine, 'citric acid', 'lightblue', "White Wine")
```


```{r}
# residual sugar
plot_density_plot(wine_data, 'residual sugar')
plot_boxplot(wine_data, 'residual sugar')
plot_boxplot_quality(wine_data, 'residual sugar')
```

```{r}
#performing normality check for red wine (residual sugar)
normality_tester(red_wine, 'residual sugar', 'red','Red Wine')

#performing normality check for white wine (residual sugar)
normality_tester(white_wine, 'residual sugar', 'lightblue', "White Wine")
```

```{r}
# chlorides
plot_density_plot(wine_data, 'chlorides')
plot_boxplot(wine_data, 'chlorides')
plot_boxplot_quality(wine_data, 'chlorides')
```

```{r}
#performing normality check for red wine (chlorides)
normality_tester(red_wine, 'chlorides', 'red','Red Wine')

#performing normality check for white wine (chlorides)
normality_tester(white_wine, 'chlorides', 'lightblue', "White Wine")
```


```{r}
# free sulfur dioxide
plot_density_plot(wine_data, 'free sulfur dioxide')
plot_boxplot(wine_data, 'free sulfur dioxide')
plot_boxplot_quality(wine_data, 'free sulfur dioxide')
```

```{r}
#performing normality check for red wine (free sulfur dioxide)
normality_tester(red_wine, 'free sulfur dioxide', 'red','Red Wine')

#performing normality check for white wine (free sulfur dioxide)
normality_tester(white_wine, 'free sulfur dioxide', 'lightblue', "White Wine")
```


```{r}
# total sulfur dioxide
plot_density_plot(wine_data, 'total sulfur dioxide')
plot_boxplot(wine_data, 'total sulfur dioxide')
plot_boxplot_quality(wine_data, 'total sulfur dioxide')
```

```{r}
#performing normality check for red wine (total sulfur dioxide)
normality_tester(red_wine, 'total sulfur dioxide', 'red','Red Wine')

#performing normality check for white wine (total sulfur dioxide)
normality_tester(white_wine, 'total sulfur dioxide', 'lightblue', "White Wine")
```


```{r}
# density
plot_density_plot(wine_data, 'density')
plot_boxplot(wine_data, 'density')
plot_boxplot_quality(wine_data, 'density')
```

```{r}
#performing normality check for red wine (density)
normality_tester(red_wine, 'density', 'red','Red Wine')

#performing normality check for white wine (density)
normality_tester(white_wine, 'density', 'lightblue', "White Wine")
```


```{r}
# pH
plot_density_plot(wine_data, 'pH')
plot_boxplot(wine_data, 'pH')
plot_boxplot_quality(wine_data, 'pH')
```

```{r}
#performing normality check for red wine (pH)
normality_tester(red_wine, 'pH', 'red','Red Wine')

#performing normality check for white wine (pH)
normality_tester(white_wine, 'pH', 'lightblue', "White Wine")
```


```{r}
# sulphates
plot_density_plot(wine_data, 'sulphates')
plot_boxplot(wine_data, 'sulphates')
plot_boxplot_quality(wine_data, 'sulphates')
```
```{r}
#performing normality check for red wine (sulphates)
normality_tester(red_wine, 'sulphates', 'red','Red Wine')

#performing normality check for white wine (sulphates)
normality_tester(white_wine, 'sulphates', 'lightblue', "White Wine")
```


```{r}
# alcohol
plot_density_plot(wine_data, 'alcohol')
plot_boxplot(wine_data, 'alcohol')
plot_boxplot_quality(wine_data, 'alcohol')

```

```{r}
#performing normality check for red wine (alcohol)
normality_tester(red_wine, 'alcohol', 'red','Red Wine')

#performing normality check for white wine (alcohol)
normality_tester(white_wine, 'alcohol', 'lightblue', "White Wine")
```

CORRELATION ANALYSIS

First we split the data into their various categories i.e continuous, ordinal and nominal categorical
Quality is ordinal categorical.
wine_type is nominal categorical.
The rest are continuous.

WINE DATA (RED AND WHITE WINES): CORRELATION MATRIX - MULTIPLE CONTINUOUS VARIABLES
```{r}
#continuous columns in the wine_data data frame
wine_data_continuous <- wine_data %>% 
  select(-quality,-wine_type)

# correlation matrix: Multiple continuous variables
# as seen in the normality tests, all the continuous variables are non-normal so we use Spearman's correlation
wine_data_continuous_cor_matrix <- round(cor(wine_data_continuous, method = "spearman"), digit=2)
wine_data_continuous_cor_matrix
```

```{r}
# Visualizing the continuous correlation matrix
corrplot(wine_data_continuous_cor_matrix,
         method = "number",
         type = "upper",
         title = "Corrplot for Continous Variables (Wine data)",
         tl.col = "black",
         number.cex = 0.6,
         tl.cex = 0.6,
         cl.cex = 0.6,
         mar = c(1, 0, 2, 0))
```
RED WINE: CORRELATION MATRIX - MULTIPLE CONTINUOUS VARIABLES
```{r}
# correlation matrix: Multiple continuous variables
red_wine_continuous <- red_wine %>% select(-quality,-wine_type)
red_wine_continuous_cor_matrix <- round(cor(red_wine_continuous, method = "spearman"), digit=2)

corrplot(red_wine_continuous_cor_matrix,
         method = "number",
         type = "upper",
         title = "Corrplot for Continous Variables (Red Wine)",
         tl.col = "red",
         number.cex = 0.6,
         tl.cex = 0.6,
         cl.cex = 0.6,
         mar = c(1, 0, 2, 0))
```
WHITE WINE: CORRELATION MATRIX - MULTIPLE CONTINUOUS VARIABLES
```{r}
# correlation matrix: Multiple continuous variables
white_wine_continuous <- white_wine %>% select(-quality,-wine_type)
white_wine_continuous_cor_matrix <- round(cor(white_wine_continuous, method = "spearman"), digit=2)

corrplot(white_wine_continuous_cor_matrix,
         method = "number",
         type = "full",
         title = "Corrplot for Continous Variables (White Wine)",
         tl.col = "blue",
         number.cex = 0.6,
         tl.cex = 0.6,
         cl.cex = 0.6,
         mar = c(1, 0, 2, 0))
```
```{r}
# Creating a copy of the wine_data df to introduce wine_type as numeric.
wine_data_copy <- wine_data %>%
  mutate(wine_type_numeric = ifelse(wine_type == "Red", 0, 1))
```


CORRELATION BETWEEN CONTINUOUS AND NOMINAL CATEGORICAL
```{r}
# Using Point-biserial correlation with wine type
point_biserial_cor <- sapply(names(wine_data_continuous), function(var) {
  cor.test(wine_data_copy[[var]], wine_data_copy$wine_type_numeric, method = "pearson")$estimate
})

# Converting to a data frame for better readability
pb_cor_df <- data.frame(
  Variable = names(wine_data_continuous),
  pb_correlation = point_biserial_cor
)
print(pb_cor_df)
```

```{r}
# Visualizing the point-biserial correlations between the continuous variables and wine type
ggplot(pb_cor_df, aes(x = reorder(Variable, pb_correlation), 
                                  y = pb_correlation, 
                                  fill = pb_correlation)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) + 
  labs(title = "Point-Biserial Correlation of Continuous Variables with Wine Type",
       x = "Continuous Variables") +
  theme_calc() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


CORREALTION BETWEEN CONTINUOUS AND ORDINAL CATEGORICAL VARAIBLES 
```{r}
# Spearman correlation of continuous variables with quality
spearman_cor <- sapply(names(wine_data_continuous), function(var) {
  cor.test(wine_data_copy[[var]], as.numeric(wine_data_copy$quality), method = "spearman")$estimate
})

# Converting results to a data frame
spearman_cor_df <- data.frame(
  Variable = names(wine_data_continuous),
  spearman_correlation = spearman_cor
)
print(spearman_cor_df)
```
```{r}
# Visualizing the spearman correlation between the continuous variables and quality
ggplot(spearman_cor_df, aes(x = reorder(Variable, spearman_correlation), 
                                  y = spearman_correlation, 
                                  fill = spearman_correlation)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient2(low = "green", mid = "white", high = "brown", midpoint = 0) + 
  labs(title = "Spearman Correlation of Continuous Variables with quality",
       x = "Continuous Variables") +
  theme_calc() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


HYPOTHESIS TESTING.
Since all the continuous variables have non-normal distributions,
We will perform 3 transformations to try to make them approximately normal
(log, sqrt, and cuberoot)

```{r}
# continuous columns in red and white wine data frames
columns_red <- names(red_wine_continuous)
columns_white <- names(white_wine_continuous)
```


```{r}
# function to log transform and perform Shapiro-Wilk test
log_shapiro <- function(data, columns) {
  p_values <- list()
  for (col in columns) {
      # log transforming
      log_transformed <- log(data[[col]])
      
      # Shapiro-Wilk test
      shapiro_test <- shapiro.test(log_transformed)
      p_values[[col]] <- shapiro_test$p.value
  }
  return(unlist(p_values))
}
```

```{r}
# function to sqrt transform and perform Shapiro-Wilk test
sqrt_shapiro <- function(data, columns) {
  p_values <- list()
  for (col in columns) {
      # sqrt transforming
      sqrt_transformed <- sqrt(data[[col]])
      
      # Shapiro-Wilk test
      shapiro_test <- shapiro.test(sqrt_transformed)
      p_values[[col]] <- shapiro_test$p.value
  }
  return(unlist(p_values))
}
```

```{r}
# function to cuberoot transform and perform Shapiro-Wilk test
cuberoot_shapiro <- function(data, columns) {
  p_values <- list()
  for (col in columns) {
      # cuberoot transforming
      sqrt_transformed <- (data[[col]])^(1/3)
      
      # Shapiro-Wilk test
      shapiro_test <- shapiro.test(sqrt_transformed)
      p_values[[col]] <- shapiro_test$p.value
  }
  return(unlist(p_values))
}
```


P-values for red wine after transforming the features
```{r}
# Dataframe showing the P-Values of the transformed variables in the red wine
red_wine_shapiro_test <- data.frame(log= log_shapiro(red_wine,columns_red),
                                    sqrt= sqrt_shapiro(red_wine,columns_red),
                                    cuberoot = cuberoot_shapiro(red_wine,columns_red))
red_wine_shapiro_test
```
P-values for white wine after transforming the features
```{r}
# Dataframe showing the P-Values of the transformed variables in the white wine
white_wine_shapiro_test <- data.frame(log= log_shapiro(white_wine,columns_white),
                                    sqrt= sqrt_shapiro(white_wine,columns_white),
                                    cuberoot = cuberoot_shapiro(white_wine,columns_white))
white_wine_shapiro_test
```
Since all the transformed variables failed to be normal based on  the Shapiro_Wilk test.
QQ plots might give us approximate normality.

```{r}
# function to plot the log transformed QQ plot
log_qqplot <- function(data, column_name, color, wine_type) {
  qqplot <- ggplot(data, aes(sample = log(.data[[column_name]]))) +
    stat_qq(color = color) +
    stat_qq_line(color = "black") +
    labs(
      title = paste("QQ Plot for log", column_name, "(", wine_type, ")"),
      x = "Theoretical",
      y = "Sample"
    ) +
    theme_calc()+ 
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
  print(qqplot)  
}
```

```{r}
# function to plot the sqrt transformed QQ plot
sqrt_qqplot <- function(data, column_name, color, wine_type) {
  qqplot <- ggplot(data, aes(sample = sqrt(.data[[column_name]]))) +
    stat_qq(color = color) +
    stat_qq_line(color = "black") +
    labs(
      title = paste("QQ Plot for sqrt", column_name, "(", wine_type, ")"),
      x = "Theoretical",
      y = "Sample"
    ) +
    theme_calc()+ 
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
  print(qqplot)  
}
```

```{r}
# function to plot the cuberoot transformed QQ plot
cuberoot_qqplot <- function(data, column_name, color, wine_type) {
  qqplot <- ggplot(data, aes(sample = (.data[[column_name]])^(1/3))) +
    stat_qq(color = color) +
    stat_qq_line(color = "black") +
    labs(
      title = paste("QQ Plot for cuberoot", column_name, "(", wine_type, ")"),
      x = "Theoretical",
      y = "Sample"
    ) +
    theme_calc()+ 
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
  print(qqplot)  
}
```

LOG TRANSFORMED QQPLOT FOR RED WINE
```{r}
# Plotting the log transformed QQ plots for all the variables in Red wine
for (i in names(red_wine_continuous)) {
  log_qqplot(red_wine, i, 'red', 'red wine')
}
```

SQRT TRANSFORMED QQPLOT FOR RED WINE
```{r}
# Plotting the sqrt transformed QQ plots for all the variables in Red wine
for (i in names(red_wine_continuous)) {
  sqrt_qqplot(red_wine, i, 'red', 'red wine')
}
```

CUBEROOT TRANSFORMED QQPLOT FOR RED WINE
```{r}
# Plotting the cuberoot transformed QQ plots for all the variables in Red wine
for (i in names(red_wine_continuous)) {
  cuberoot_qqplot(red_wine, i, 'red', 'red wine')
}
```

LOG TRANSFORMED QQPLOT FOR WHITE WINE
```{r}
# Plotting the log transformed QQ plots for all the variables in White wine
for (i in names(white_wine_continuous)) {
  log_qqplot(white_wine, i, 'lightblue', 'white wine')
}
```
SQRT TRANSFORMED QQPLOT FOR WHITE WINE
```{r}
# Plotting the sqrt transformed QQ plots for all the variables in White wine
for (i in names(white_wine_continuous)) {
  sqrt_qqplot(white_wine, i, 'lightblue', 'white wine')
}
```


CUBEROOT TRANSFORMED QQPLOT FOR WHITE WINE
```{r}
# Plotting the cuberoot transformed QQ plots for all the variables in White wine
for (i in names(white_wine_continuous)) {
  cuberoot_qqplot(white_wine, i, 'lightblue', 'white wine')
}
```


Hypotheses Testing 1

Tests if there is an association between the wine type and quality

Since wine_type is nominal categorical and quality is ordinal

Chi-Square Test of Independence

Hypotheses:
Null Hypothesis (H₀):There is no association between wine type and quality
Alternative Hypothesis (H₁): Wine type and wine quality are associated
```{r}
# Creating quality type to reduce the perform Chi-squared test
wine_data_copy <- wine_data_copy %>%
  mutate(quality_type = case_when(
    quality %in% c(3, 4) ~ "Low",
    quality %in% c(5,6) ~ "Medium",
    quality %in% c(7,8, 9) ~ "High"
  ))
wine_data_copy$quality_type  <- factor(wine_data_copy$quality_type , 
                                   levels = c("Low", "Medium", "High"), 
                                   ordered = TRUE)
table(wine_data_copy$wine_type,wine_data_copy$quality_type)
```

```{r}
# Plotting number of wines by Quality and type
ggplot(wine_data_copy,aes(x = quality_type, fill = wine_type)) +
  geom_bar() +
  labs(title = "Number of Wines by Quality by Type") +
  scale_fill_manual(values = c("Red" = "red", "White" = "lightblue")) +  theme_calc()
```


```{r}
# Chi-squared test of independence.
chisq_test <- chisq.test(wine_data_copy$wine_type, wine_data_copy$quality_type)
chisq_test
chisq_test$residuals
```
Since the p-value is much less than 0.05, we reject the null hypothesis. 
This means there is a statistically significant association between wine_type and quality_type.

High-quality wines: Fewer for red wine, more for white wine.
Medium-quality wines: Slightly more for red wine, slightly fewer for white wine.


HYPOTHESIS TEST 2.

Testing pH vs Wine Type (Two Sample T-Test)

Null Hypothesis (H₀) : There is no significant difference in the mean pH between the two wine types.
Alternative Hypothesis (H₁):There is significant difference in the mean pH between the two wine types

```{r}
# Visualizing the distribution of log_pH based on wine type.
wine_data %>% 
  mutate(log_pH = log(pH)) %>% 
  ggplot(aes(x = wine_type, y = log_pH, fill = wine_type)) +
    geom_boxplot() +
    labs(title = paste("Boxplot of log_pH by Wine Type")) +
    scale_fill_manual(values = c("Red" = "red", "White" = "lightblue")) +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))

wine_data %>% 
  mutate(log_pH = log(pH)) %>% 
  ggplot(aes(x = log_pH, fill = wine_type)) +
  geom_density(alpha = 0.7) +
    labs(title = paste("Density plot of log_pH by Wine Type")) +
    scale_fill_manual(values = c("Red" = "red", "White" = "lightblue")) +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
```


ASSUMPTIONS

Normality check
```{r}
# plotting histograms to check normality distribution of log_pH
red_wine %>%
  mutate(log_pH = log(pH)) %>% 
  ggplot(aes(x = log_pH)) +
    geom_histogram(alpha = 0.7, bins =10,fill = "red") +
    labs(title ="Histogram of log_pH (Red Wine)") +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))

white_wine %>%
  mutate(log_pH = log(pH)) %>% 
  ggplot(aes(x = log_pH)) +
    geom_histogram(alpha = 0.7, bins =10, fill = "lightblue") +
    labs(title ="Histogram of log_pH (White Wine)") +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
```

```{r}
# Shapiro-Wilk test for normality
wine_data %>%
  group_by(wine_type) %>%
  mutate(log_pH = log(pH)) %>% 
  summarise(p_value = shapiro.test(log_pH)$p.value)
```
P<0.05, w ereject null hyposthesis of normality

```{r}
# QQ plot check for normality
red_wine %>%
  mutate(log_pH = log(pH)) %>%
  ggplot(aes(sample = log_pH)) +
    stat_qq(color ='red') +
    stat_qq_line(color = "black") +
    labs(
      title = paste("QQ Plot for log_pH of red wine"),
      x = "Theoretical",
      y = "Sample") +
    theme_calc()+ 
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))

white_wine %>%
  mutate(log_pH = log(pH)) %>%
  ggplot(aes(sample = log_pH)) +
    stat_qq(color = 'lightblue') +
    stat_qq_line(color = "black") +
    labs(
      title = paste("QQ Plot for log_pH of white wine"),
      x = "Theoretical",
      y = "Sample") +
    theme_calc()+ 
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
```
QQ plots appear approximately normal. We can carry on with a parametric test.

```{r}
# Test for homogeneity of variance
wine_data %>% 
  mutate(log_pH = log(pH)) %>% 
  bartlett.test(log_pH ~ wine_type, data = .)
```
P>0,05, fail to reject null hypothesis, variance is equal.

```{r}
# Two sample t-test (two tailed)
wine_data %>% 
  mutate(log_pH = log(pH)) %>% 
  t.test(log_pH ~ wine_type, data = ., var.equal = TRUE)
```
In this case the p-value is less than 0.05, therefore the Null Hypothesis is rejected and Alternative Hypothesis is accepted.



Hypothesis Test 3

chlorides vs. Wine Type
Hypotheses:
Null Hypothesis (H₀) The distribution of chlorides is identical in red and white wines.
Alternative Hypothesis (H₁):The distribution of chlorides is not identical in red and white wines.

```{r}
#normality checks
# Red wine
shapiro.test(red_wine$chlorides)
# White wine
shapiro.test(white_wine$chlorides)
```

p<0.05: Not normally distributed → Mann-Whitney U Test.

Assumptions for Mann-Whitney U Test
Checking for same shape
```{r}
#plotting the distribution of chlorides for red  wine
  ggplot(red_wine, aes(x = chlorides)) +
    geom_histogram(alpha = 0.7, bins =10,fill = "red") +
    labs(title ="Histogram of chlorides (Red Wine)") +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
#plotting the distribution of chlorides for white wine
  ggplot(white_wine, aes(x = chlorides)) +
    geom_histogram(alpha = 0.7, bins =10, fill = "lightblue") +
    labs(title ="Histogram of chlorides (White Wine)") +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
```

```{r}
# Mann-Whitney U Test / Wilcox rank sum test
wilcox.test(chlorides ~ wine_type,data = wine_data)
```
p<0.05: we reject Null Hypothesis (H₀):The chlorides distributions differ significantly between the two wine types

Hypothesis Test 4
Hypotheses:
Null Hypothesis (H₀): Median Alcohol level is equal across all quality type
Alternative Hypothesis (H₁): Median Alcohol is not equal across all quality type


```{r}
# Normality check for alcohol for all quality types
byf.shapiro(alcohol ~ quality_type, data = wine_data_copy)
```
P<0.05, Reject null hypothesis. Not normally distributed.

```{r}
# Test for homogeneity of variance
bartlett.test(alcohol ~ quality_type, data = wine_data_copy)
```
P > 0.05, Accept Null hypothesis.

```{r}
# Distribution of alcohol in across quality type
ggplot(data=wine_data_copy, aes(x = quality_type, y = alcohol, fill=quality_type)) +
  geom_boxplot() +
  labs(title = paste("Alcohol level for each Quality Type")) +
  scale_fill_manual(values = c("Low" = "lightgreen", "Medium" = "green", "High" = "darkgreen")) +
  theme_calc() +
  theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
```


```{r}
#plotting the distribution of Alcohol for low quality
wine_data_copy %>% 
  filter(quality_type == 'Low') %>% 
  ggplot(aes(x = alcohol)) +
    geom_density(fill = 'lightgreen') +
    labs(title ="Desnity Plot of Alcohol for Low Quality") +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
```

```{r}
#plotting the distribution of Alcohol for medium quality
wine_data_copy %>% 
  filter(quality_type == 'Medium') %>% 
  ggplot(aes(x = alcohol)) +
    geom_density(fill = 'green') +
    labs(title ="Desnity Plot of Alcohol for Medium Quality") +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
```

```{r}
#plotting the distribution of Alcohol for high quality
wine_data_copy %>% 
  filter(quality_type == 'High') %>% 
  ggplot(aes(x = alcohol)) +
    geom_density(fill = 'darkgreen') +
    labs(title ="Desnity Plot of Alcohol for High Quality") +
    theme_calc() +
    theme(plot.title = element_text(size = 14, margin = margin(b = 10)))
```

```{r}
# Kruskal-Wallace test for non parametric models.
kruskal.test(alcohol ~ quality_type, data = wine_data_copy)
```
p < 0.05, We reject Null hypothesis and carry out Dunes test, to check which is actually different.

POST-HOC TEST
```{r}
# Dunns test 
dunnTest(wine_data_copy$alcohol, wine_data_copy$quality_type, method = 'bonferroni')
```



Regression


Model 1
Question: Can wine type and physiochemical properties be used to predict wine quality.
Logistic Regression 1

```{r}
# preparing data for logistic regression
# changing the quality to binary for logistic regression
wine_data_logistic <- wine_data %>%
  mutate(quality_binary = case_when(
    quality %in% c(3, 4,5,6) ~ 0,
    quality %in% c(7,8, 9) ~ 1
  ))
head(wine_data_logistic)
```
Backward Step wise
```{r}
# Quailty prediction model 1
quality_prediction_1 <- glm(
  quality_binary ~ wine_type + `fixed acidity` + `volatile acidity` +
  `citric acid`+ `residual sugar` + chlorides+ `free sulfur dioxide` +
  `total sulfur dioxide`+ sulphates + density + pH + alcohol,
  data = wine_data_logistic,family = "binomial"
)

summary(quality_prediction_1)
```

```{r}
# Quality prediction model 2
quality_prediction_2 <- glm(
  quality_binary ~ wine_type + `fixed acidity` + `volatile acidity` +
  `residual sugar` + chlorides+ `free sulfur dioxide` +
  `total sulfur dioxide`+ sulphates + density + pH + alcohol,
  data = wine_data_logistic,family = "binomial"
)

summary(quality_prediction_2)
```
```{r}
# Important features for predicting quality
quality_feature_importance <- varImp(quality_prediction_2, scale=False)
quality_feature_importance <- quality_feature_importance %>% arrange(desc(Overall))
quality_feature_importance
```
Most important features when predicting quality are pH, volatile acidity,sulphates,residual sugar,fixed acidity
density,alcohol.
```{r}
# Using only the most important features
quality_prediction_3 <- glm(
  quality_binary ~ `fixed acidity` + `volatile acidity` +`residual sugar`+
  sulphates + density + pH + alcohol,
  data = wine_data_logistic,family = "binomial"
)

summary(quality_prediction_3)
```

```{r}
# Multicolinearity check
vif(quality_prediction_2)
```

```{r}
# Quality prediction model 4
quality_prediction_4 <- glm(
  quality_binary ~ `volatile acidity` + wine_type +
  `residual sugar` + chlorides+ `free sulfur dioxide`+ sulphates + pH + alcohol,
  data = wine_data_logistic,family = "binomial"
)

summary(quality_prediction_4)
```
```{r}
# Multicollinearity check
vif(quality_prediction_4)
```

Linearity Check
```{r}
# Calculating pi
probs_quality <- predict(quality_prediction_4, data=wine_data_logistic,type="response")
wine_data_logistic$probs_quality <- probs_quality
```

```{r}
# Calculating logit(pi)
wine_data_logistic$logits_quality <- log(probs_quality/(1-probs_quality))
```


```{r}
# scatter plot for linearity check
pairs(wine_data_logistic[,c(16,2,9,11)], lower.panel = NULL,
      upper.panel = panel.smooth, pch = 19,cex = 0.2)
```
Logits appears to have a linear relationship with volatile acidity, pH and alcohol.

```{r}
# Cooks disatnce test for influential values
plot(quality_prediction_4, which = 4, id.n = 3)
```
It is worth noting that observation 246,935, and 5187, might have outliers.


Model 2
Question: Can wine quality and physiochemical properties be used to predict wine type?

Logistic Regression Model 2

```{r}
# data for prediction (removing columns observed with substantial outliers)
wine_data_logistic_2 <- wine_data %>%
  mutate(wine_type_numeric = ifelse(wine_type == "Red", 0, 1)) %>% 
  slice(-c(3654,919,929))
```


Backward Step wise
```{r}
# First Prediction model for wine type
type_prediction_1 <- glm(
  wine_type_numeric ~ quality + `fixed acidity` + `volatile acidity` +
  `citric acid`+ `residual sugar` + chlorides+ `free sulfur dioxide` +
  `total sulfur dioxide`+ sulphates + density + pH + alcohol,
  data = wine_data_logistic_2,family = "binomial"
)

summary(type_prediction_1)
```

```{r}
# Second prediction model for wine type.
type_prediction_2 <- glm(
  wine_type_numeric ~ `volatile acidity` +
  `residual sugar`+ `total sulfur dioxide`+ density + alcohol,
  data = wine_data_logistic_2,family = "binomial"
)

summary(type_prediction_2)
```

```{r}
# Most important variables when predicting wine type
type_feature_importance <- varImp(type_prediction_2, scale=False)
type_feature_importance <- type_feature_importance %>% arrange(desc(Overall))
type_feature_importance
```
Most important features when predicting wine type are density, total sulfur dioxide, residual sugar, volatile acidity and alcohol

```{r}
# Multicolinearity check
vif(type_prediction_2)
```

Linearity check
```{r}
# Calculating pi values
probs_type <- predict(type_prediction_2, data=wine_data_logistic_2,type="response")
wine_data_logistic_2$probs_type <- probs_type
```

```{r}
# calculating (logit(pi))
wine_data_logistic_2$logits_type <- log(probs_type/(1-probs_type))
```


```{r}
# Scatter plot to confirm linearity
pairs(wine_data_logistic_2[,c(16,8,4,11)], lower.panel = NULL,
      upper.panel = panel.smooth, pch = 19,cex = 0.2)
```
Logits has an approximately linear relationship with density,residual sugar and alcohol.

```{r}
# checking for influential values.
plot(type_prediction_2, which = 4, id.n = 3)
```
Noticed that at observation 920,2752 and 4516, have some influential values worth looking out for.

```{r}
# Regression curve to predict wine type.
ggplot(wine_data_logistic_2, aes(x=`volatile acidity` +
  `residual sugar`+ `total sulfur dioxide`+ density + alcohol,
  y=wine_type_numeric)) + 
geom_point() + 
stat_smooth(method="glm", color="lightgreen", se=FALSE, 
method.args = list(family=binomial))

```





MODEL 3
Question: can the other physiochemical properties be used to predict the density of wines?
Multiple Linear Regression model.

```{r}
# Visualizing the relationship between density and other properties
corrplot(wine_data_continuous_cor_matrix,
         method = "circle",
         type = "upper",
         title = "Relationship Between the Physiochemical Properties",
         tl.col = "black",
         number.cex = 0.6,
         tl.cex = 0.6,
         cl.cex = 0.6,
         mar = c(1, 0, 2, 0))
```
It seems like Density has a good correlation with alcohol, fixed acidity, volatile acidity and residual sugar.

```{r}
# removing outliers using Z score
compute_z_scores <- function(data) {
  as.data.frame(scale(data))}
```

```{r}
# Removing outliers
z_scores <- compute_z_scores(wine_data_continuous)
outlier_rows <- apply(z_scores, 1, function(row) any(abs(row) > 3))
wine_no_outliers <- wine_data_continuous[!outlier_rows, ]
```

```{r}
#Multiple Linear regression model
density_prediction <- lm(density ~ alcohol +`residual sugar` +
                        `fixed acidity`+ `volatile acidity`,
                         wine_no_outliers)
summary(density_prediction)
```


```{r}
# Linearity between the IVs and DV
pairs(wine_no_outliers[,c(8,1,4,11)], lower.panel = NULL,
      upper.panel = panel.smooth,pch = 19,cex = 0.2)
```

```{r}
# Linearity between the IVs and DV
crPlots(density_prediction)
```
All the independent variables have a linear relationship with density (Dependent variable)

```{r}
# Residual Independence
plot(density_prediction,1)
```
The correlation between the residuals is approximately 0, therefore theresiduals are independent.

```{r}
# Normality of residuals
plot(density_prediction,2)
```
The residuals appear approximately normally distributed, with some deviations at the tail

```{r}
# Homoscedacity test
plot(density_prediction,3)
```
There is no clear pattern among residuals, they appear randomly scattered with equal variablilty.

```{r}
# Multicollinearity
vif(density_prediction)
```
The VIF of all the IVs are all less than 2, which indicates little to no correlation between the IVs and DV.


